How to Avoid OOTA Without Really Trying

### 2024-09-12 paper/ootavacuous.tex
ootavacuous: Apply more Alan Stern September 12 feedback
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

In broad terms, OOTA occurs when a group of threads load from each others'
stores and each thread's store depends on the value returned by that
thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air''
as shown in
Listing~\ref{lst:Simple OOTA}.
Here, all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have final
values of 42, despite there being no instances of 42 in the default-zero
initial values or in the executable
code~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.\footnote{
	Throughout this paper, variables whose names are ``\co{r}''
	followed by a number denote CPU registers, and other names denote
	shared variables.
	All shared-variable accesses are relaxed atomic.}
First, P0's line~1 does a load from \co{x} into \co{r1},
reading the value of P1's line-2 store rather than \co{x}'s initial
value of zero and somehow obtaining 42.
Second, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, P1's line~1 loads 42 from \co{y} to \co{r2}.
Finally, P1's line~2 stores 42 to \co{x}, justifying the value loaded
by P0's line~1.

Because nothing else in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly prohibits them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950},
which prevents misapplication of the as-if rule in oracular C++
implementations:
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
This paper uses \emph{full C++} to denote the standard including this
prohibition of OOTA and \emph{loose C++} to denote a hypothetical standard
that excludes this prohibition but is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

On the other hand, OOTA cycles need not involve nonsensical or
unmotivated values, as shown in
Listing~\ref{lst:OOTA Cycle}.
Line~3 of both processes stores the value 42 only when line~2 determines
that the value loaded was 42.
However, this OOTA result has not been observed on real implementations
due to the default-zero initialization combined with semantic dependencies
from line~1 to line~3.

>>>>

In broad terms, OOTA occurs when a group of threads load from each others'
stores and each thread's store depends on the value returned by that
thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air''
as shown in
Listing~\ref{lst:Simple OOTA}.
Here, all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have final
values of 42, despite there being no instances of 42 in the default-zero
initial values or in the executable
code~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.\footnote{
	Throughout this paper, variables whose names are ``\co{r}''
	followed by a number denote private per-process variables
	(``registers''),
	and other names denote shared variables.
	All shared-variable accesses are relaxed atomic.}
First, process P0's line~1 does a load from \co{x} into \co{r1},
reading the value of P1's line-2 store rather than \co{x}'s initial
value of zero and somehow obtaining 42.
Second, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, P1's line~1 loads 42 from \co{y} to \co{r2}.
Finally, P1's line~2 stores 42 to \co{x}, justifying the value loaded
by P0's line~1.

Because nothing else in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly prohibits them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}
(which prevents misapplication of the as-if rule in oracular C++
implementations):
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
This paper uses \emph{full C++} to denote the standard including this
prohibition of OOTA and \emph{loose C++} to denote a hypothetical standard
that excludes this prohibition but is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

On the other hand, OOTA cycles need not involve nonsensical or
unmotivated values, as shown in
Listing~\ref{lst:OOTA Cycle}.
Line~3 of both processes stores the value 42 only when line~2 determines
that the value loaded was 42.
This, combined with the default-zero initialization, explains why this
OOTA result has not been observed on real implementations.
